Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package

### 2021-06-22 010-background.Rmd
Revise for JOSS submission. Address Ed's comments.
Revise for JOSS submission. Address Walt's comments.

\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, definitely by \@ref(eq:inverseCDF) when the variables are discrete.

@@

To place `Bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.  In `R`, the `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks. Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD `R` packages exist, such as `GenOrd`. For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table \@ref(tab:compare-table) summarizes the above discussion.

>>>>

\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, defined by \@ref(eq:inverseCDF) when the variables are discrete.

@@

To place `Bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.  In `R`, the `MASS` package's `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks. Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD `R` packages exist, such as `GenOrd`. For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table \@ref(tab:compare-table) summarizes the above discussion.

### 2021-06-22 020-simulation-algorithm.Rmd
Revise for JOSS submission. Address Walt's comments.

1. Mapping step, either
    * Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
    * Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
    * Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.
2. Check admissibility and, if needed, compute the nearest correlation matrix.
    * Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.
    * If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise,
    * Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.
3. Gaussian copula
    * Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.
    * Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
    * Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.

>>>>

1. Mapping step, either
    * Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
    * Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
    * Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.
2. Check correlation matrix admissibility and, if not, compute the nearest correlation matrix.
    * Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.
    * If the mapping produced an $R_X$ that is a correlation matrix, we describe the matrix as **admissible** in this scheme and retain $R_X$. Otherwise,
    * Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm and the result will be approximate.
3. Gaussian copula
    * Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.
    * Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
    * Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.

### 2021-06-22 030-bigsimr-package.Rmd
Revise for JOSS submission. Address Walt's comments.

The subsections below describe the basic use of `bigsimr` R package by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983]. This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.

@@

Finally, we execute `rvec` to simulate the desired $10,000$ random vectors from the assumed bivariate distribution of `Ozone` and `Temp`. Note that the target input in specified using the pre-computed MVN correlation matrix. Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.

>>>>

The subsections below describe the basic use of the `bigsimr` R package by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983]. This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.

@@

Finally, we execute `rvec` to simulate the desired $10,000$ random vectors from the assumed bivariate distribution of `Ozone` and `Temp`. Note that the target input is specified using the pre-computed MVN correlation matrix. Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.

### 2021-06-22 050-examples.Rmd
Revise for JOSS submission. Address Walt's comments.

Next, we estimate the marginal parameters. We use method of moments to estimate the marginal parameters for the multivariate negative binomial model. The marginal distributions are from the same probability family (NB), yet they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$. The functions below support this estimation for later use in `rvec`.

@@

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig). Figure \@ref(fig:ch050-figBRCA) compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis). The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlations from the data.

>>>>

Next, we estimate the marginal parameters. We use the method of moments to estimate the marginal parameters for the multivariate negative binomial model. The marginal distributions are from the same probability family (NB), yet they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$. The functions below support this estimation for later use in `rvec`.

@@

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points than the observed data in Figure \@ref(fig:ch010-realDataFig). Figure \@ref(fig:ch050-figBRCA) compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis). The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlations from the data.

### 2021-06-22 080-appendix.Rmd
Revise for JSS submission pre cross ref fix
Revise for JOSS submission. Address Walt's comments.

\noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal). First, let us note that when we get such a vector from the `bigsimr` package then in fact it can be represented as

@@

\begin{proof}
The values of the means and the variances are obtained immediately from normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$. It remains to establish the covariance of $Y_1$ and $Y_2$,

@@

Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This approximately agrees with the MC results provided.

>>>>

\noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal). First, let us note that when we generate such a vector from the `bigsimr` package then in fact it can be represented as

@@

\begin{proof}
The values of the means and the variances are obtained immediately from the normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$. It remains to establish the covariance of $Y_1$ and $Y_2$,

@@

Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This agrees with the MC results provided.

### 2021-06-22 _book/_main.tex
Revise for JOSS submission. Address Walt's comments.

\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, definitely by \eqref{eq:inverseCDF} when the variables are discrete.

@@

To place \texttt{Bigsimr}'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages. In \texttt{R}, the \texttt{MASS} package \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions. This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy \texttt{random.multivariate\_normal} both do similar tasks. Recently, the high-performance \texttt{R} package \texttt{mvnfast} \citep{Fasiolo2016} has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD \texttt{R} packages exist, such as \texttt{GenOrd}. For copula computation, the full-featured \texttt{R} package \texttt{copula} provides many different copulas and a fairly general LD random generator \texttt{rCopula()}. High-dimesional random vectors via \texttt{copula} takes an impractical amount of computing time \citep{Li2019gpu}. The \texttt{nortaRA} R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other \texttt{R} package that can produce HD random vectors with flexible margins and dependence modeling. Table \ref{tab:compare-table} summarizes the above discussion.

@@

  \begin{itemize}
  \tightlist
  \item
    Convert the target Spearman correlation matrix \(R_S\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
  \end{itemize}
\item
  Check admissibility and, if needed, compute the nearest correlation matrix.

  \begin{itemize}
  \tightlist
  \item
    Check that \(R_X\) is a correlation matrix, a positive definite matrix with 1's along the diagonal.
  \item
    If \(R_X\) is a correlation matrix, the input matrix is \emph{admissible} in this scheme. Otherwise,
  \item
    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm.
  \end{itemize}
\item
  Gaussian copula

>>>>

\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, defined by \eqref{eq:inverseCDF} when the variables are discrete.

@@

To place \texttt{Bigsimr}'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages. In \texttt{R}, the \texttt{MASS} package's \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions. This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy \texttt{random.multivariate\_normal} both do similar tasks. Recently, the high-performance \texttt{R} package \texttt{mvnfast} \citep{Fasiolo2016} has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD \texttt{R} packages exist, such as \texttt{GenOrd}. For copula computation, the full-featured \texttt{R} package \texttt{copula} provides many different copulas and a fairly general LD random generator \texttt{rCopula()}. High-dimesional random vectors via \texttt{copula} takes an impractical amount of computing time \citep{Li2019gpu}. The \texttt{nortaRA} R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other \texttt{R} package that can produce HD random vectors with flexible margins and dependence modeling. Table \ref{tab:compare-table} summarizes the above discussion.

@@

  \begin{itemize}
  \tightlist
  \item
    Convert the target Spearman correlation matrix \(R_S\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
  \end{itemize}
\item
  Check correlation matrix admissibility and, if not, compute the nearest correlation matrix.

  \begin{itemize}
  \tightlist
  \item
    Check that \(R_X\) is a correlation matrix, a positive definite matrix with 1's along the diagonal.
  \item
    If the mapping produced an \(R_X\) that is a correlation matrix, we describe the matrix as \textbf{admissible} in this scheme and retain \(R_X\). Otherwise,
  \item
    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm and the result will be approximate.
  \end{itemize}
\item
  Gaussian copula
